\scuba: iterative map-making with the Sub-Millimetre User Reduction Facility

### 2012-11-13 smurf_paper.tex
Improve caption for Fig.9
addressing mostly smaller comments in 1st half of paper

\begin{figure*}
\includegraphics[width=0.49\linewidth]{bolos_point_mix_s2sro}
\includegraphics[width=0.49\linewidth]{bolos_point_mix}
\caption{A comparison between single bolometer time-series in each of
  the 450 and 850\,\micron\ bands with the mixing chamber temperature
  and azimuth/elevation pointing offsets, before and after upgrading
  the instrument. The grey signals over-plotted in the top two panels
  show the residual time-series after removing the common-mode
  signals.  Left: data taken during the S2SRO period, observation 29
  on 20100313. There is a strong correlation between the bolometers
  and the roughly $\sim$25\,s oscillation in the fridge, but only a
  minor correlation with the telescope motion. The nearly flat
  common-mode subtracted signals show: (i) that most of the
  low-frequency signal is common to all of the bolometers; and (ii)
  the non-correlated, and predominantly white noise at 450\,\micron\
  is significantly larger than at 850\,\micron. Right: data taken with
  the fully-commissioned instrument, observation 38 on
  20111112. Unlike the S2SRO data, there is no strong signal produced
  by variations in the fridge temperature. The 450\,\micron\ data have
  minimal correlated low-frequency noise (as evidenced by the lack of
  a difference between the raw and common-mode subtracted signals),
  although the 850\,\micron\ data show a common signal correlated with
  the telescope motion (note that this scan has a significantly larger
  amplitude than the S2SRO data set), mostly likely caused by magnetic
  field pickup (see Section~\ref{sec:magpickup} and
  Fig.~\ref{fig:magpickup}).}
\label{fig:bolos_mix}
\end{figure*}

@@

The majority of the correlated signal at both wavelengths in
Fig.~\ref{fig:pca} is clearly produced by Components~1 and 2. In both
cases, the eigenvector time-series exhibit a roughly periodic signal
that resembles the scan pattern in the right-hand panel of
Fig.~\ref{fig:pca}. While these correlated signals also clearly appear
at both wavelengths, referring to the maps of eigenvalues, they are
considerably stronger at 850\,\micron\ (consistent with the visual
appearance of the bolometer signals in Fig.~\ref{fig:pca}). Also note
that the PSDs for these eigenvectors exhibit nearly a pure $1/f$
signature with almost no line features, or a white-noise plateau
(suggesting that these are high-\snr\ measurements of a purely
low-frequency drift).

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{dimm}
\caption{Typical map-making algorithm. Raw data (stored in multiple
  files) are read and concatenated into continuous time-series, and
  flatfieded. Based on the scan speed of the telescope, the data are
  down-sampled to match the selected output map pixel size (usually 2
  and 4\,arcsec at 450 and 850\,\micron, respectively, to ensure
  adequate sampling of the \scuba\ PSFs). A cleaning stage repairs DC
  steps and spikes, and fills gaps to ensure continuity. Finally, the
  mean of each bolometer time-series is removed (higher-order
  polynomials may also be used). The iterative portion then begins
  (dashed box): estimating and removing the common-mode signal (the
  combined \model{COM} and \model{GAI} models); applying the
  extinction correction (\model{EXT}); high-pass filtering to remove
  residual independent low-frequency noise (\model{FLT}); estimating
  the map by re-gridding the data, and then removing its projection
  from the time-series (\model{AST}); and finally measuring the noise
  in the residual time-series (\model{NOI}). If the solution has
  converged, the map is written to disk. Otherwise any multiplicative
  factors that may have been applied to the data are inverted (i.e.,
  the extinction correction, \model{EXT}), and then the models for the
  low-frequency noise; the common-mode (\model{COM,GAI}) and high-pass
  filter (\model{FLT}). Each model is then re-estimated in turn until
  \model{AST}, at which point the previous estimate of the
  astronomical signal is added back into the data prior to its
  re-calculation.}
\label{fig:dimm}
\end{figure}

First, the raw data are passed through a pre-processing stage which
corrects some of the more significant glitches, applies flat-field
corrections etc. Next, the iterative process begins (dashed box). Most
of the low-frequency noise is removed using common-mode subtraction
(\model{COM,GAI}, Section~\ref{sec:comgai}). Then, the extinction
correction (\model{EXT}, Section~\ref{sec:ext}) is applied. At this
point the data resemble the grey bolometer traces in
Fig.~\ref{fig:bolos_mix}, with some residual baseline drifts still
visible. These drifts are removed using a high-pass filter
(\model{FLT}, Section~\ref{sec:flt}) implemented with FFTs. The
residual signal is then re-gridded to estimate the map. Finally, the
map is used to estimate and remove the astronomical signal
contamination in the bolometer data (\model{AST},
Section~\ref{sec:ast}), leaving a relatively clean data set in which
to measure the white noise level of each bolometer (\model{NOI},
Section~\ref{sec:noi}). Since the common-mode and filtering stages
will have introduced ringing in the map in the vicinity of bright
sources, the entire process is iterated. Each component in the dashed
box of Fig.~\ref{fig:dimm} is re-calculated in sequence. In this way,
the second time the common-mode is calculated, for example, most of
the bright astronomical sources in the data have already been removed
in the previous iteration when the map was estimated, reducing the
amount of ringing in the map once it is re-estimated. Note that the
extinction correction is a multiplicative factor, and must be inverted
prior to re-calculating any of the additive model components to
preserve the data amplitude. Additive model components from the
previous iteration are generically added back into the time-series
immediately prior to their re-calculation. However, for the special
case of the common-mode and high-pass filter stages, they are replaced
simultaneously at the start of the iteration to assist with
convergence (see Sections~\ref{sec:converge}).

@@

The method used by SMURF is to average together multiple samples from
the original time-series, $x_i$ to estimate the lower-frequency output
time-series, $y_j$. In general there will not be an integer number of
samples from $x_i$ in each of the $y_j$, so fractional samples from
$x_i$ are used at the $y_j$ time boundaries. The algorithm is fast,
since each sample in $x_i$ need only be visited once. From a spectral
point of view, this boxcar average is equivalent to applying a
sinc-function low-pass filter. Such behaviour is desirable, since it
serves as an anti-aliasing filter; most non-white features above the
Nyquist frequency in $y_j$ that could contaminate the output data are
removed.

@@

SMURF uses FFTs of the bolometer data extensively for filtering.  Data
that have been flagged as bad for any reason (for instance, due to the
presence of spikes, steps, or unusual common-mode signal) need to be
excluded from this FFT. For this reason, each contiguous block of bad
data samples is filled with artificial data before taking the FFT. A
least squares linear fit is performed to the 50 samples preceding the
block, and a similar fit is performed to the 50 samples following the
block. These are used to estimate the expected values at the start and
end of the block of bad values. The bad values are then replaced by
linear interpolation between the expected start and end
values. Gaussian noise is added with a standard deviation equal to the
mean of the RMS residuals in the two fits.

@@

First, we describe the generic model for \scuba\ data. We express the
digitised signal observed by the $i$th bolometer as a function of
time,
%
\begin{equation}
\mathbf{b}_i(t) = f_i[\mathbf{e}_i(t) \mathbf{a}_i(t) + \mathbf{n}_i(t)],
\label{eq:model}
\end{equation}
%
where $\mathbf{a}(t)$ is the time-varying signal produced by scanning
the telescope across astronomical sources, $\mathbf{e}(t)$ is the
time-varying extinction, which is a function of the telescope
elevation and atmospheric conditions, and $\mathbf{n}_i(t)$ represents
sources of noise. The two terms in square brackets, as written, have
units of power delivered to the detectors (pW). The scale factor $f_i$
converts this effective power to the digitised units recorded on disk
(DAC) -- the flatfield multiplied by a digitisation constant (applied
once during pre-processing) -- which in this formulation is assumed to
be constant in time.

@@

During pre-processing, the map-maker divides by $f_i$ once. Then, in
each iteration, the map-maker divides by $f_i$, models and removes
$g_i\mathbf{n}^\mathrm{c}(t)$, divides by $\mathbf{e}_i(t)$, and
applies a high-pass filter to remove $\mathbf{n}^\mathrm{f}_i(t)$ from
the bolometer time-series. This procedure isolates the astronomical
signal and white noise, $\mathbf{a}_i(t) +
\mathbf{n}^\mathrm{w}_i(t)$, for estimating the map. Finally, once the
map is estimated it is projected into the time-domain and removed from
the bolometer time-series leaving $\mathbf{n}^\mathrm{w}_i(t)$, in
which the bolometer noise can be measured. This is the most typical
sequence; in practice the user can select an arbitrary order, although
the solution converges much faster when the strongest components are
estimated first, and then subtracted, leaving a cleaner signal for
subsequent model estimates.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{com}
\caption{A selection of typical bolometer time-series after
flatfielding and removal of a constant baseline. It can be seen that
most bolometers exhibit a common time variation overlaid with other
features. Empty squares indicate the locations of broken bolometers.}
\label{fig:com}
\end{figure}

@@

It can be seen from Fig.~\ref{fig:com} that some bolometers depart
radically from the common-mode, indicating some problem with the
bolometer. Such bolometers are identified by calculating the Pearson
correlation coefficient between each bolometer time-series and the
common mode. Bolometers for which the correlation coefficient is below
a specified limit, or which have unusually high or low gains compared
to the other bolometers, are flagged as bad in order to omit them from
the final map. If the option described above for handling time-varying
sensitivities is used, then a correlation coefficient can be
determined for each individual block of time slices.  This allows
individual bad blocks to be rejected from a bolometer time-series,
rather than rejecting the whole bolometer. As a warning, rejecting
data based on outlier gain values can be misleading in cases where the
data are dominated by magnetic field pickup. For example, the
common-mode signal for the data in Fig.~\ref{fig:magpickup} would
resemble the azimuthal scan pattern, but clearly the gains will have
opposite signs (such that one or the other bolometer would be
erroneously flagged).

@@

Any astronomical sources that are smaller than the array size will
contribute signal to some bolometers but not other bolometers, thus
biassing the simple mean used to estimate the common mode. However, on
each iteration of the map-making algorithm illustrated in
Fig.~\ref{fig:dimm}, a large fraction of the remaining astronomical
signal is extracted from the bolometer time-series and transferred to the
output map, resulting in subsequent estimates of the common-mode being
more accurate.

@@

Map estimation is accomplished using a nearest-neighbour resampling of
the data onto a pre-defined map grid. For the $i$th map pixel,
$\mathbf{m}(x_i,y_i)$, the brightness is estimated as the weighted
average of the bolometer data samples $\mathbf{b}_j$ that land within
that pixel (from any bolometer or point in time),
%
\begin{equation}
  \mathbf{m}(x_i,y_i) = \frac{\sum_j \mathbf{w}_j \mathbf{b}_j }
                             { \sum_j \mathbf{w}_j } .
\end{equation}
%
For the initial iteration the weights $\mathbf{w}_j$ are set to 1, but
subsequently they are set to $1/\sigma_j^2$, the estimated inverse
variance expected from the bolometer white noise levels, as discussed
in Section~\ref{sec:noi}. This weighting scheme is sensible provided
that the bolometer data have no correlated (e.g., low-frequency)
noise.

In addition to the signal map, a variance map $\mathbf{v}(x_j,y_j)$ is
estimated. The default procedure is to estimate this quantity given
the scatter in the weighted samples. This is accomplished by dividing
the biased weighted sample variance by the number of samples that went
into the average (akin to the formula for standard error on the mean,
but accounting for weights),
%
\begin{equation}
\label{eq:varmap}
\mathbf{v}(x_j,y_j) = \frac{\sum_j \mathbf{w}_j
                            \sum_j \mathbf{w}_j \mathbf{b}_j -
                            \left( \sum_j \mathbf{w}_j \mathbf{b}_j \right)^2 }
                           { N \left( \sum_j \mathbf{w}_j \right)^2 },
\end{equation}
%
where $N_j$ is the total number of bolometer samples that land in the
pixel. As written, this algorithm is numerically unstable if the two
terms in the numerator are large, and of nearly the same value:
floating point precision errors can cause the difference to be
significantly incorrect. In practice we use the more numerically
stable ``weighted incremental algorithm'' that calculates incremental
differences, as described in \citet{west1979}.

@@

In addition to map estimation, the \model{AST} model can also be used
to perform map-based despiking of the time-series. Unlike the
time-domain despiker (Section~\ref{sec:timedespike}), this calculation
utilises the scatter in the population of samples used to estimate the
map pixel values (from different times and bolometers) to reject
outliers. This method is more robust against false-positive detections
of bright/compact astronomical sources, since transient features in
the time-series are unlikely to occur by chance whenever a bolometer
crosses a specific location on the sky, whereas real astronomical
sources are at fixed spatial locations.

@@

Since noise estimation is usually calculated as the final step in the
iteration, the data at this stage have had most of the astronomical
and other large, low-frequency noise signals removed. For this reason,
\model{NOI} may optionally perform some cleaning options, such as the
DC step fixer (Section~\ref{sec:steps}) and spike detection
(Section~\ref{sec:timedespike}), which may work better with these
cleaner time-series.

@@

\subsubsection{Other experimental models}

@@

As described in Section~\ref{sec:magpickup}, wide scans can produce
significant magnetic field pickup that tracks the azimuthal motion of
the telescope. The \model{DKS} model uses the dark squid signals for
each column as a template that is simply scaled (gain and offset) to
each bolometer time-series before removal. Since not all columns have
working dark squids, another alternative model is \model{TMP} in which
the azimuth of the telescope itself is used as the template. In both
cases, while the models had some success at removing the magnetic
field pickup, they did not reduce the $1/f$ knee substantially,
necessitating the continued use of the \model{FLT} model to remove the
remaining low-frequency noise. Little or no difference in the final
maps could be seen with or without the inclusion of these models, and
so they are not generally used.

@@

Finally, we experimented with an alternative to \model{COM} for
removing correlated atmospheric noise. Rather than subtracting the
average signal at each time slice, \model{PLN} fits a plane to the
observed signal at each instant. We found that there was no obvious
improvement (either in terms of reducing the $1/f$ knee or reducing
the noise in final maps). This result is basically consistent with
those from earlier instruments, and suggests that the angular scale of
the emitting regions in the atmosphere are un-resolved at the \scuba\
focal plane. \citet{sayers2010}, in particular, examined several
different ways of modelling and removing the correlated atmospheric
noise from Bolocam data at 143 and 268\,GHz (also atop Mauna Kea),
finding only marginal improvements by fitting a plane, or even using
higher-order polynomials (see their figure~10, as well as the
discussion and references to earlier work in their section~4.3). In
\citet{aguirre2011} iterative PCA cleaning was used to remove the
atmosphere since the simpler methods mentioned left substantial
residuals in their data (similar to the \scuba\ data described here,
although PCA cleaning is prohibitively slow in our case).

@@

The map-maker will halt if the convergence criteria have been
met. Presently there are two numerical tests that may be performed by
SMURF, or alternatively, a fixed number of iterations may be
specified. The first numerical test is an approximate check of the
change in reduced chi-squared, $\chi^2_\mathrm{r}$. The standard
deviations, $\sigma_i$, are measured for each of the residual
bolometer signals once the modelled signal components have been
removed. In the early iterations, the residual signals usually contain
both white noise, and other long-timescale features. However, once the
solution has converged, this signal should look approximately
white. Once the white noise, $\mathbf{n}^\mathrm{w}_i(t)$, has been
measured from the bolometer PSDs in \model{NOI}
(Section~\ref{sec:noi}), $\chi^2_\mathrm{r}$ is calculated as $(1/N)
\sum_i \sigma^2_i / \mathbf{n}^\mathrm{w}_i(t)$, where $i$ runs over
the $N$ bolometers. In other words, it is the average ratio between
the measured time-series bolometer variances and their white-noise
levels measured between 2 and 10\,Hz. Note that this expression does
not account for the degrees of freedom. Clearly there are a large
number of parameters in the model, which will ``fit-out'' some of the
uncorrelated white noise, and potentially bias this estimate of
$\chi^2_\mathrm{r}$ low. One approach would be to keep track of the
degrees of freedom, and therefore provide a correction, as in
\citet{kovacs2008}. However, note that we are only using this quantity
as a convergence test, and its value is not important in an absolute
sense. Furthermore, the reference white noise values are calculated
only once in SMURF (Section~\ref{sec:noi}), and the number of model
parameters are fixed, so the value of $\chi^2_\mathrm{r}$ should
decrease monotonically if the solution is behaving well. The
convergence criterion is met if the change in subsequent values of
$\chi^2_\mathrm{r}$ is smaller than the requsted threshikd, which is
set to $10^{-3}$ by default.

>>>>

\begin{figure*}
\centering
\includegraphics[width=0.49\linewidth]{bolos_point_mix_s2sro}
\includegraphics[width=0.49\linewidth]{bolos_point_mix}
\caption{A comparison between single bolometer time-series in each of
  the 450 and 850\,\micron\ bands with the mixing chamber temperature,
  the line-of-sight 450\,\micron\ opacity derived from the water
  vapour monitor, and azimuth/elevation pointing offsets, before and
  after upgrading the instrument. The grey signals over-plotted in the
  top two panels show the residual time-series after removing the
  common-mode signals.  Left: data taken during the S2SRO period,
  observation 29 on 20100313. There is a strong correlation between
  the bolometers and the roughly $\sim$25\,s oscillation in the
  fridge, but only a minor correlation with the opacity and telescope
  motion. The nearly flat common-mode subtracted signals show: (i)
  that most of the low-frequency signal is common to all of the
  bolometers; and (ii) the non-correlated, and predominantly white
  noise at 450\,\micron\ is significantly larger than at
  850\,\micron. Right: data taken with the fully-commissioned
  instrument, observation 38 on 20111112. Unlike the S2SRO data, there
  is no strong signal produced by variations in the fridge
  temperature. The 450\,\micron\ data have minimal correlated
  low-frequency noise (as evidenced by the lack of a difference
  between the raw and common-mode subtracted signals), although the
  850\,\micron\ data show a common signal correlated with the opacity
  and telescope motion (note that this scan has a significantly larger
  amplitude than the S2SRO data set), mostly likely caused by a
  combination of changing airmass (anti-correlated with opacity), and
  magnetic field pickup (see Section~\ref{sec:magpickup} and
  Fig.~\ref{fig:magpickup}).}
\label{fig:bolos_mix}
\end{figure*}

@@

The majority of the correlated signal at both wavelengths in
Fig.~\ref{fig:pca} is clearly produced by Components~1 and 2. In both
cases, the eigenvector time-series exhibit a roughly periodic signal
that resembles the scan pattern in the right-hand panel of
Fig.~\ref{fig:bolos_mix}. While these correlated signals also clearly
appear at both wavelengths, referring to the maps of eigenvalues, they
are considerably stronger at 850\,\micron\ (consistent with the visual
appearance of the bolometer signals in Fig.~\ref{fig:pca}). Also note
that the PSDs for these eigenvectors exhibit nearly a pure $1/f$
signature with almost no line features, or a white-noise plateau
(suggesting that these are high-\snr\ measurements of a purely
low-frequency drift).

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{dimm}
\caption{Typical map-making algorithm. Raw data (stored in multiple
  files) are read and concatenated into continuous time-series, and
  flatfieded. Based on the scan speed of the telescope, the data are
  down-sampled to match the selected output map pixel size (usually 2
  and 4\,arcsec at 450 and 850\,\micron, respectively, to ensure
  adequate sampling of the \scuba\ PSFs). A cleaning stage repairs
  steps and spikes, and fills gaps to ensure continuity. Finally, the
  mean of each bolometer time-series is removed (higher-order
  polynomials may also be used). The iterative portion then begins
  (dashed box): estimating and removing the common-mode signal (the
  combined \model{COM} and \model{GAI} models); applying the
  extinction correction (\model{EXT}); high-pass filtering to remove
  residual independent low-frequency noise (\model{FLT}); estimating
  the map by re-gridding the data, and then removing its projection
  from the time-series (\model{AST}); and finally measuring the noise
  in the residual time-series (\model{NOI}). If the solution has
  converged, the map is written to disk. Otherwise any multiplicative
  factors that may have been applied to the data are inverted (i.e.,
  the extinction correction, \model{EXT}), and then the models for the
  low-frequency noise; the common-mode (\model{COM,GAI}) and high-pass
  filter (\model{FLT}). Each model is then re-estimated in turn until
  \model{AST}, at which point the previous estimate of the
  astronomical signal is added back into the data prior to its
  re-calculation.}
\label{fig:dimm}
\end{figure}

First, the raw data are passed through a pre-processing stage which
corrects some of the more significant glitches, applies flat-field
corrections etc. Next, the iterative process begins (dashed box). Most
of the low-frequency noise is removed using common-mode subtraction
(\model{COM,GAI}, Section~\ref{sec:comgai}). Then, the extinction
correction (\model{EXT}, Section~\ref{sec:ext}) is applied. At this
point the data resemble the grey bolometer traces in
Fig.~\ref{fig:bolos_mix}, with some residual baseline drifts still
visible. These drifts are removed using a high-pass filter
(\model{FLT}, Section~\ref{sec:flt}) implemented with FFTs. The
residual signal is then re-gridded to estimate the map. Finally, the
map is used to estimate and remove the astronomical signal from the
bolometer data (\model{AST}, Section~\ref{sec:ast}), leaving a data
set that is appropriate for measuring the white noise level of each
bolometer (\model{NOI}, Section~\ref{sec:noi}). Since the common-mode
and filtering stages will have introduced ringing in the map in the
vicinity of bright sources, the entire process is iterated. Each
component in the dashed box of Fig.~\ref{fig:dimm} is re-calculated in
sequence. In this way, the second time the common-mode is calculated,
for example, most of the bright astronomical sources in the data have
already been removed in the previous iteration when the map was
estimated, reducing the amount of ringing in the map once it is
re-estimated. Note that the extinction correction is a multiplicative
factor, and must be inverted prior to re-calculating any of the
additive model components to preserve the data amplitude. Additive
model components from the previous iteration are generically added
back into the time-series immediately prior to their
re-calculation. However, for the special case of the common-mode and
high-pass filter stages, they are replaced simultaneously at the start
of the iteration to assist with convergence (see
Sections~\ref{sec:converge}).

@@

The method used by SMURF is to average together multiple samples from
the original time-series, $x_i$ to estimate the lower-frequency output
time-series, $y_j$. In general there will not be an integer number of
samples from $x_i$ in each of the $y_j$, so fractional samples from
$x_i$ are used at the $y_j$ time boundaries: $y_j = (\sum w_i
x_i)/(\sum w_i)$, where $i$ runs over all the samples from $x$ that
land within, or partially overlap, $y_j$, and the weights,
$0<w_i\le1$, indicate the fractions of each $x_i$ that land within
$y_j$. The algorithm is fast since each sample in $x_i$ need only be
visited once. From a spectral point of view, this boxcar average is
equivalent to applying a sinc-function low-pass filter. Such behaviour
is desirable, since it serves as an anti-aliasing filter; most
non-white features above the Nyquist frequency in $y_j$ that could
contaminate the output data are removed.

@@

SMURF uses FFTs of the bolometer data extensively for filtering. Data
that have been flagged as bad for any reason (for instance, due to the
presence of spikes, steps, or unusual common-mode signal) when need to
be excluded from this FFT. For this reason, each contiguous block of
bad data samples is filled with artificial data before taking the
FFT. A least squares linear fit is performed to the 50 samples
preceding the block, and a similar fit is performed to the 50 samples
following the block. These are used to estimate the expected values at
the start and end of the block of bad values. The bad values are then
replaced by linear interpolation between the expected start and end
values. Gaussian noise is added with a standard deviation equal to the
mean of the RMS residuals in the two fits. These flagged and filled
portions of the data are then given a weight of zero when estimating
the map.

@@

First, we describe the basic model for \scuba\ data. We express the
digitised signal observed by the $i$th bolometer as a function of
time,
%
\begin{equation}
\mathbf{b}_i(t) = f_i[\mathbf{e}_i(t) \mathbf{a}_i(t) + \mathbf{n}_i(t)],
\label{eq:model}
\end{equation}
%
where $\mathbf{a}(t)$ is the time-varying signal produced by scanning
the telescope across astronomical sources, $\mathbf{e}(t)$ is the
time-varying extinction, which is a function of the telescope
elevation and atmospheric conditions, and $\mathbf{n}_i(t)$ represents
sources of noise. The two terms in square brackets, as written, have
units of power delivered to the detectors (pW). The scale factor $f_i$
converts this effective power to the digitised units recorded on disk
(DAC) -- the flatfield multiplied by a digitisation constant (applied
once during pre-processing) -- which in this formulation is assumed to
be constant in time.

@@

During pre-processing, the map-maker divides by $f_i$ once. Then, in
each iteration, the map-maker models and removes
$g_i\mathbf{n}^\mathrm{c}(t)$, divides by $\mathbf{e}_i(t)$, and
applies a high-pass filter to remove $\mathbf{n}^\mathrm{f}_i(t)$ from
the bolometer time-series. This procedure isolates the astronomical
signal and white noise, $\mathbf{a}_i(t) +
\mathbf{n}^\mathrm{w}_i(t)$, for estimating the map. Finally, once the
map is estimated it is projected into the time-domain and removed from
the bolometer time-series leaving $\mathbf{n}^\mathrm{w}_i(t)$, in
which the bolometer noise can be measured. This is the most typical
sequence; in practice the user can select an arbitrary order, although
the solution converges much faster when the strongest components are
estimated first, and then subtracted, leaving a cleaner signal for
subsequent model estimates.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{com}
\caption{A selection of typical bolometer time-series after
flatfielding and removal of a constant baseline. Each sub-plot shows 
the power in pW from a single bolometer over a single 30 second period.  
It can be seen that most bolometers exhibit a common time variation overlaid with other
features. Empty squares indicate the locations of broken bolometers.}
\label{fig:com}
\end{figure}

@@

It can be seen from Fig.~\ref{fig:com} that some bolometers depart
radically from the common-mode, indicating some problem with the
bolometer. Such bolometers are identified by calculating the Pearson
correlation coefficient between each bolometer time-series and the
common mode. Bolometers for which the correlation coefficient is below
a specified limit, or which have unusually high or low gains compared
to the other bolometers, are flagged as bad in order to omit them from
the final map. If the option described above for handling time-varying
sensitivities is used, then a correlation coefficient can be
determined for each individual block of time slices.  This allows
individual bad blocks to be rejected from a bolometer time-series,
rather than rejecting the whole bolometer. As a warning, rejecting
data based on outlier gain values can be misleading in cases where the
data are dominated by magnetic field pickup. For example, the
common-mode signal for the data in Fig.~\ref{fig:magpickup} would
resemble the azimuthal scan pattern, but clearly the gains will have
opposite signs (such that one or the other bolometer would be
erroneously flagged). Experimental models that were investigated for
magnetic field pickup removal are described in
Section~\ref{sec:experimental}.

@@

Any astronomical sources that are smaller than the array size will
contribute signal to some bolometers but not other bolometers, thus
biassing the simple mean used to estimate the common mode. However, on
each iteration of the map-making algorithm illustrated in
Fig.~\ref{fig:dimm}, a large fraction of the remaining astronomical
signal is extracted from the bolometer time-series and transferred to
the output map, resulting in subsequent estimates of the common-mode
being more accurate. It is also for this reason that the previously
described step of (re-)flagging outlier portions of the data is
generally performed iteratively, rather than once as a pre-processing
step, since bright/compact astronomical sources can be the cause of
rejection in early iterations.

@@

Map estimation is accomplished using a nearest-neighbour resampling of
the data onto a pre-defined map grid. For the $i$th map pixel,
$\mathbf{m}(x_i,y_i)$, the brightness is estimated as the weighted
average of the bolometer data samples $\mathbf{b}_j$ that land within
that pixel (from any bolometer or point in time, provided that they
are not flagged as bad or gap-filled, Section~\ref{sec:gaps}),
%
\begin{equation}
  \mathbf{m}(x_i,y_i) = \frac{\sum_j \mathbf{w}_j \mathbf{b}_j }
                             { \sum_j \mathbf{w}_j } .
\end{equation}
%
For the initial iteration the weights $\mathbf{w}_j$ are set to 1, but
subsequently they are set to $1/\sigma_j^2$, the estimated inverse
variance expected from the bolometer white noise levels, as discussed
in Section~\ref{sec:noi}. This weighting scheme is sensible provided
that the bolometer data have no correlated (e.g., low-frequency)
noise.

In addition to the signal map, a variance map $\mathbf{v}(x_j,y_j)$ is
estimated. The default procedure is to estimate this quantity given
the scatter in the weighted samples. This is accomplished by dividing
the biased weighted sample variance by the number of samples that went
into the average (akin to the formula for standard error on the mean,
but accounting for weights),
%
\begin{equation}
\label{eq:varmap}
\mathbf{v}(x_j,y_j) = \frac{\sum_j \mathbf{w}_j
                            \sum_j \mathbf{w}_j \mathbf{b}_j -
                            \left( \sum_j \mathbf{w}_j \mathbf{b}_j \right)^2 }
                           { N \left( \sum_j \mathbf{w}_j \right)^2 },
\end{equation}
%
where $N_j$ is the total number of bolometer samples that land in the
pixel. As written, this algorithm is numerically unstable if the two
terms in the numerator are large, and of nearly the same value:
floating point precision errors can cause the difference to be
significantly incorrect. In practice we use the superior ``weighted
incremental algorithm'' that calculates incremental differences, as
described in \citet{west1979}.

@@

In addition to map estimation, the \model{AST} model can also be used
to perform map-based despiking of the time-series. Unlike the
time-domain despiker (Section~\ref{sec:timedespike}), this calculation
utilises the scatter in the population of samples that land in a map
pixel (from different times and bolometers) to reject outliers. This
method is more robust against false-positive detections of
bright/compact astronomical sources since transient features in the
time-series are unlikely to occur by chance whenever a bolometer
crosses a specific location on the sky, whereas real astronomical
sources are at fixed spatial locations.

@@

Since noise estimation is usually calculated as the final step in the
iteration, the data at this stage have had most of the astronomical
and other large, low-frequency noise signals removed. For this reason,
\model{NOI} may optionally perform some cleaning options, such as the
step fixer (Section~\ref{sec:steps}) and spike detection
(Section~\ref{sec:timedespike}), which may work better with these
cleaner time-series.

@@

\subsubsection{Experimental models for the removal of magnetic field pickup and the atmosphere}
\label{sec:experimental}

@@

As described in Section~\ref{sec:magpickup}, wide scans can produce
significant magnetic field pickup that tracks the azimuthal motion of
the telescope. The \model{DKS} model uses the dark squid signals for
each column as a template that is simply scaled (gain and offset) to
each bolometer time-series before removal. Unfortunately the dark
squids do not work for every column (several in each subarray),
meaning that those columns are usually discarded when producing
maps. One possible remedy is to identify dead bolometers (with
otherwise working TES readouts) or intentionally disconnect working
bolometers, to create replacement dark readouts in those columns. This
solution has not been pursued due to the limited success at removing
pickup using the presently working dark squids. Another alternative
model is \model{TMP} in which the azimuth of the telescope itself is
used as the template. For both \model{DKS} and \model{TMP}, there is
some success at visibly decreasing the low-frequency scan-synchronous
noise in certain data sets, although often it does not (and there are
also examples in which the noise is increased). In the case of
\model{DKS}, there may simply be components of the signal seen by
working detectors that are not detecting in the dark
squids. Furthermore, in the case of \model{TMP}, the relationship
between the projection of the Earth's magnetic field on the instrument
and the magnitude of the pickup may not be linearly correlated as we
have assumed (verified in some cases by a comparison of the azimuthal
motion with working dark squid signals). In no case was the use of
these models able to reduce the $1/f$ knee substantially,
necessitating the continued use of \model{FLT} to remove the remaining
low-frequency noise. Neither \model{DKS} nor \model{TMP} are typically
used.

@@

Finally, we experimented with an alternative to \model{COM} for
removing correlated atmospheric noise. Rather than subtracting the
average signal at each time slice, \model{PLN} fits a plane to the
observed signal at each instant. We found that there was no obvious
improvement (either in terms of reducing the $1/f$ knee or reducing
the noise in final maps). This result is basically consistent with
those from earlier instruments \citep[e.g., the prediction for \scuba\
based on SCUBA data described in][]{chapin2002}, and suggests that the
angular scale of the emitting regions in the atmosphere are
un-resolved at the \scuba\ focal plane. \citet{sayers2010}, in
particular, examined several different ways of modelling and removing
the correlated atmospheric noise from Bolocam data at 143 and 268\,GHz
(also atop Mauna Kea), finding only marginal improvements by fitting a
plane, or even higher-order polynomials (see their figure~10, as well
as the discussion and references to earlier work in their
section~4.3). In \citet{aguirre2011} iterative PCA cleaning was used
to remove the atmosphere since the simpler methods mentioned left
substantial residuals in their data (similar to the \scuba\ data
described here, although PCA cleaning is prohibitively slow in our
case).

@@

The map-maker will halt if the convergence criteria have been
met. Presently there are two numerical tests that may be performed by
SMURF, or alternatively, a fixed number of iterations may be
specified. The first numerical test is an approximate check of the
change in reduced chi-squared, $\chi^2_\mathrm{r}$. The standard
deviations, $\sigma_i$, are measured for each of the residual
bolometer signals once the modelled signal components have been
removed. In the early iterations, the residual signals usually contain
both white noise, and other long-timescale features. However, once the
solution has converged, this signal should look approximately
white. Once the white noise, $\mathbf{n}^\mathrm{w}_i(t)$, has been
measured from the bolometer PSDs in \model{NOI}
(Section~\ref{sec:noi}), $\chi^2_\mathrm{r}$ is calculated as $(1/N)
\sum_i \sigma^2_i / \mathbf{n}^\mathrm{w}_i(t)$, where $i$ runs over
the $N$ bolometers. In other words, it is the average ratio between
the measured time-series bolometer variances and their white-noise
levels measured between 2 and 10\,Hz. Note that this expression does
not account for the degrees of freedom. Clearly there are a large
number of parameters in the model, which will ``fit-out'' some of the
uncorrelated white noise, and potentially bias this estimate of
$\chi^2_\mathrm{r}$ low. One approach would be to keep track of the
degrees of freedom, and therefore provide a correction, as in
\citet{kovacs2008}. However, note that we are only using this quantity
as a convergence test, and its value is not important in an absolute
sense. Furthermore, the reference white noise values are calculated
only once in SMURF (Section~\ref{sec:noi}), and the number of model
parameters are fixed, so the value of $\chi^2_\mathrm{r}$ should
decrease monotonically if the solution is behaving well. The
convergence criterion is met if the change in subsequent values of
$\chi^2_\mathrm{r}$ is smaller than the requsted threshold, which is
set to $10^{-3}$ by default.
